Exploring the Role of Small Molecules in Biological Systems Using Network Approaches

### 2015-03-02 paper.tex
Added section on scaffold doc network

The traditional, reductionist view of biology has considered the
behavior and interactions of individual molecular entities as a way to
understand and explain biological phenomena. This approach has become
embodied in the ``one-target, one-drug''  approach to drug
development. Leaving aside the subtleties of how targets are defined,
it has become apparent that this model does not correspond very well
to reality. Instead, it has become increasingly clear that many drugs
(and more generally, small molecules) interact with multiple targets,
including their putative primary target.

@@

@@

>>>>

The traditional, reductionist view of biology has considered the
behavior and interactions of individual molecular entities as a way to
understand and explain biological phenomena. This approach has become
embodied in the ``one-target, one-drug'' approach to drug
development. Leaving aside the subtleties of how targets are defined,
it has become apparent that this model does not always match
reality. Instead, it has become increasingly clear that many drugs
(and more generally, small molecules) exhibit polypharmacological
behavior - interacting with multiple targets, including their putative
primary target. In many cases, the activity against the primary target
is significantly higher than off-target activities, minimizing the
effects of the latter. However, this is not always the case as in the
example of terfenadine (a histamine H1 antagonist), where off target
activity against hERG led to an increased risk of cardiac death. But
more importantly, it is becoming increasingly clear that many diseases
are not driven by dysfunction of a single a target. Examples include
diabetes and various cancers. In parallel, even when a single target
is the main driver for a given disease state, compensatory pathways
can ``route around'' inhibition of the primary target, thus restoring
the functional behavior that the primary target is responsible (i.e.,
the system is now resistant to the original drug)
\cite{Sierra:2010mz,Dienstmann:2012qr}. Finally, it should be noted
that while the preceding discussion presents polypharmacological
behavior as a negative characteristic, this is not always the
case. Indeed, modern approaches to drug development now consider such
behavior attractive leading to efforts to rationally design or control
polypharmacology \cite{Gujral:2014dp,Ciceri:2014fv,Metz:2010ys}.

Given these observations it is clear that pathways play fundamental
roles in biological systems. Naturally, approaches that analyse these
networked systems can be effective in understanding their behavior and
then identifying ways to modulate them. The remainder of this chapter
looks at how network methods can be applied to various topics in drug
discovery ranging from drug-target interactions to analyzing
structure-activity relationships.

@@

\subsection{Scaffold-document networks}
\label{sec:scaff-docum-netw}

With the advent of SAR databases such as ChEMBL \cite{Gaulton:2012jl},
it is feasible to explore chemical space from a bibliometric point of
view. For example, ChEMBL has identified small molecules mentioned in
publications and provides the links between them. As a result, given a
set of document it is possible to characterize the chemical space
represented in those documents. As noted in Section
\ref{sec:scaffold-networks}, scaffolds (or frameworks) allow one to
better capture the core structures representing multiple
compounds. Thus, a publication might refer to ten compounds from a
single series, in which case it refers to a single scaffold (though
this could depend on how one defines a scaffold). To explore this
aspect of a SAR database, we applied a fragment decomposition method
to all compounds reported in ChEMBL and linked the fragments
associated with each molecule to their corresponding documents. The
resultant data can then be represented as a bipartite network of
fragments and documents which are connected when a document contains
molecules that have the fragment as a substructure.

One application of this dataset is to explore the nature of fragments
associated with a document, and for each of those fragments identify
the documents associated with them. This procedure can be repeated to
construct scaffold-document network that represents the chemical and
bibliometric space associated with a starting document. As an example
of this type of analysis we constructed a network starting from a
report describing a series of imidazonaphthalimides and their
antitumor activity \cite{Brana:2002pr}. The network was limited to 3
iterations (document $\rightarrow$ fragment $\rightarrow$ document
$\rightarrow$ fragment). In addition we applied constraints on the
fragments (e.g., must be contained in 20 to 50 molecules and exhibit a
minimum complexity), resulting in a network with 1277 nodes (816
document nodes and 461 fragment nodes) shown in Figure
\ref{fig:scafodcnet}. 
\begin{figure}[h]
  \centering
  \includegraphics[width=\linewidth]{img/12477366}
  \caption{An example of a scaffold-document network, limited to three
    iterations. Squares represent fragments and circles represent
    documents. Document nodes are colored based on their major topic
    MeSH heading. The seed document is shown with a red border (top
    left).}
  \label{fig:scafdocnet}
\end{figure}
In this figure, fragment nodes are squares and document nodes are
circles, colored by the major topic MeSH heading. (Only the 12 most
frequent MeSH headings are considered - document nodes outside this
set are colored grey). Though difficult to navigate as a static  
figure, the network structure is interesting in that the immediate
neighborhood of the seed document (red border, top left) is relatively
small. The median number of fragments associated with a document is 1,
(though two documents were connected to more than 100
fragments). Conversely, the median number of documents connected to a
fragment was 3. Fragments tend to be associated with multiple
documents (which is expected based on the number of iterations used to
construct the network, as well as the constraints on fragment
characteristics) and based on the MeSH headings there is topical
overlap between many of these documents. For example, most document
clusters are annotated with the \emph{Chemistry} and \emph{Chemical
  synthesis} MeSH headings. Given that the seed document has been
annotated with \emph{Chemical synthesis} it is not surprising that
based on the chemistry contained within it, related documents will
also be related to synthesis. Importantly, if one considers all
documents in ChEMBL (with valid Pubmed ID's), the two most frequent
MeSH headings are \emph{Chemistry} and \emph{Chemical synthesis} and
thus overrepresentation of these terms is unsurprising.

This discussion has focused on the document nodes in this network. One
could also examine the fragment nodes, in terms of their similarity to
each other. Alternatively, one could examine the ``history'' of a
fragment in terms of its representation in publications over
time. Since ChEMBL also provides bioactivity data for the compounds in
a document this network representation could be used to track the
progression of bioactivity of a scaffold (or rather its members) over
time.

@@

% While the \rcdk package and its dependencies provide general small
% molecule handling functionality, a number of other packages provide
% more specific functionality, related to small molecules. We have
% previously mentioned DrugComboRanker \cite{huang2014drugcomboranker}
% and MANTRA \cite{iorio2010discovery}.
